Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-11-06 inference.tex
edits

The rely-guarantee framework presented in the previous section
facilitates modular proofs for weakly-isolated transactions, but
imposes a non-trivial annotation burden.  In particular, it requires
each statement ($c$) of the transaction to be annotated with a stable
pre- ($P$) and post-condition ($Q$), and loops to be annotated with
stable inductive invariants ($\psi$). While weakest pre-condition
style predicate transformers can help in inferring intermediate
assertions for regular statements, loop invariant inference remains
challenging, even for the simple form of loops considered here.  As an
alternative, we present an inference algorithm based on state
transformers that alleviates this burden.  The idea is to infer the
logical effect that each statement has on the transaction-local
database state $\stl$ (i.e., how it transforms $\stl$), and compose
multiple such effects together to describe the effect of the
transaction as a whole.  Importantly, this approach generalizes to
loops, where the effect of a loop can be inferred as a well-defined
function of the effect of its body, thanks to certain pleasant
properties enjoyed by the database programs modeled by $\txnimp$.
Interpreting database semantics as functional transformations on sets
(described in terms of their logical effects) enables an inference
mechanism that can leverage off-the-shelf SMT solvers for automated
verification.

@@

At the core of our approach is a simple language ($\SL$) to express
set transformations (see Fig.~\ref{fig:logic-syntax}). The language
admits only set expressions that include variables ($x$), literals of
the form $\{x \,|\, \varphi\}$ where $\varphi$ is a propositional
(quantifier-free) formula on $x$, a restricted form of existential
quantification that binds a set $\stg$ satisfying proposition $\phi$ in a
set expression $s$, a monadic composition of two set expressions
($s_1$ and $s_2$) composed using a bind ($\bind$) operation, a
conditional set expression where the condition is a propositional
formula, and a union of two set expressions. Symbols $\stl$ and $\stg$
are also variables in $\SL$, but are used to denote local and database
states, respectively. Constant sets can be written using set
literal expressions. For example, the set $\{1,2\}$ can be written as
$\{x \,|\, x=1 \vee x=2\}$. The language is carefully chosen to be
expressive enough to capture the semantics of $\txnimp$ statements (as
well as SQL operations more generally), yet simple enough to have a
semantics-preserving translation amenable for automated verification.

@@

In the above RG judgment, let us call the pre-condition $P$, and
post-condition $Q$. The stability condition on the state transformer

@@

$\Fx(\stg) \cup \F(\stg)$, where $\stg$ and $\stg')$ are related by

@@

The proofs for \C{SELECT}, sequencing, and conditionals use use the
inductive hypothesis to infer the RG-judgments present in the
premises of their corresponding RG-rules. The interesting case is the
\C{FOREACH} statement, for which we use the loop invariant
$\psi(\stl, \stg) \Leftrightarrow \stl = s \cup (y \bind (\lambda
z.~\F(s, \stg)))$, (where assuming that $c$ is the body of the loop,
$c \elabsto \F$) to prove all the premises of
\rulelabel{RG-Foreach}. Using the same notation as the rule
$\rulelabel{RG-Foreach}$, $y$ refers to the records already processed
in previous iterations of the loop, while $z$ refers to the record
being processed in the current iteration.  At the beginning of the
loop $[\phi/y]\psi(\stl, \stg)$ just reduces to $\stl = s$ which is
implied by the pre-condition $P$. From the inductive hypothesis, we
can infer that each iteration corresponds to the application of
$\F$. Since all iterations are assumed to be independent of each other,
and $z$ is bound to a record in $x$ for each iteration, we conclude
that at the end of every iteration, the loop invariant $[y \cup
  \{z\}/y]\psi$ will be satisfied.

>>>>

The rely-guarantee framework presented in the previous section
facilitates modular proofs for weakly-isolated transactions, but
imposes a non-trivial annotation burden.  In particular, it requires
each statement ($c$) of the transaction to be annotated with a stable
pre- ($P$) and post-condition ($Q$), and loops to be annotated with
stable inductive invariants ($\psi$). While weakest pre-condition
style predicate transformers can help in inferring intermediate
assertions for regular statements, loop invariant inference remains
challenging, even for the simple form of loops considered here.  As an
alternative, we present an inference algorithm based on state
transformers that alleviates this burden.  The idea is to infer the
logical effect that each statement has on the transaction-local
database state $\stl$ (i.e., how it transforms $\stl$), and compose
multiple such effects together to describe the effect of the
transaction as a whole.  Importantly, this approach generalizes to
loops, where the effect of a loop can be inferred as a well-defined
function of the effect of its body, thanks to certain pleasant
properties enjoyed by the database programs modeled by our core
language.  Interpreting database semantics as functional
transformations on sets (described in terms of their logical effects)
enables an inference mechanism that can leverage off-the-shelf SMT
solvers for automated verification.

@@

At the core of our approach is a simple language ($\SL$) to express
set transformations (see Fig.~\ref{fig:logic-syntax}). The language
admits set expressions that include variables ($x$), literals of the
form $\{x \,|\, \varphi\}$ where $\varphi$ is a propositional
(quantifier-free) formula on $x$, a restricted form of existential
quantification that binds a set $\stg$ satisfying proposition $\phi$
in a set expression $s$, a monadic composition of two set expressions
($s_1$ and $s_2$) composed using a bind ($\bind$) operation, a
conditional set expression where the condition is a propositional
formula, and a union of two set expressions. Symbols $\stl$ and $\stg$
are also variables in $\SL$, but are used to denote local and database
states (also represented as sets), respectively. Constant sets can be
written using set literal expressions. For example, the set $\{1,2\}$
can be written as $\{x \,|\, x=1 \vee x=2\}$. The language is
carefully chosen to be expressive enough to capture the semantics of
$\txnimp$ statements (as well as SQL operations more generally), yet
simple enough to have a semantics-preserving translation amenable for
automated verification.

@@

In the above RG judgment, let us call the pre-condition
($\hoare{\lambda(\stl,\stg).~ \stl = \Fx(\stg)}$) $P$, and
post-condition (${\lambda(\stl,\stg).~ \stl = \Fx(\stg) \cup
  \F(\stg)}$), $Q$. The stability condition on the state transformer

@@

$\Fx(\stg) \cup \F(\stg)$, where $\stg$ and $\stg'$ are related by

@@

The proofs for \C{SELECT}, sequencing, and conditionals use the
inductive hypothesis to infer the RG-judgments present in the
premises of their corresponding RG-rules. The interesting case is the
\C{FOREACH} statement, for which we use the loop invariant
$\psi(\stl, \stg) \Leftrightarrow \stl = s \cup (y \bind (\lambda
z.~\F(s, \stg)))$, (where assuming that $c$ is the body of the loop,
$c \elabsto \F$) to prove all the premises of
\rulelabel{RG-Foreach}. Using the same notation as the rule
$\rulelabel{RG-Foreach}$, $y$ refers to the records already processed
in previous iterations of the loop, while $z$ refers to the record
being processed in the current iteration.  At the beginning of the
loop $[\phi/y]\psi(\stl, \stg)$ just reduces to $\stl = s$ which is
implied by the pre-condition $P$. From the inductive hypothesis, we
can infer that each iteration corresponds to the application of
$\F$. Since all iterations are assumed to be independent of each other,
and $z$ is bound to a record in $x$ for each iteration, we conclude
that at the end of every iteration, the loop invariant $[y \cup
  \{z\}/y]\psi$ will be satisfied.
